Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2025-07-20 _manuscript/revision/response.qmd

---
title: "Response to Reviewers JDS24112"
subtitle: "Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks"
author: "H. Sherry Zhang, Roger D. Peng"
date: "2025-07-08"
format: html
bibliography: ../references.bib
---

@@

We would like to thank the reviewers and editor for taking the time to review our paper and provide constructive comments. Below is a point-by-point explanation of how we have changed the paper. Reviewers' comments are in normal text and **our responses are in bold text**.

# Reviewer 1

## General comments

1. The manuscript is being considered for the special issue on Statistical Aspects of Trustworthy Machine Learning. However, the connection to the theme of the special issue feels somewhat underdeveloped. It would be helpful for the authors to clarify which aspect(s) of trustworthiness are being addressed. While the work appears to relate to robustness and reproducibility, these connections should be more clearly articulated and substantiated in the manuscript. **A: We now discuss the connection between our work and the trustworthiness of machine learning in the third paragraph of the *Introduction*. Specifically, our work address the breakdown in trust caused by  violations of algorithmic assumptions -- corresponding to the third step in @broderick_toward_2023. In modern data analysis, where data are often not collected directly by the analysts, assumptions about the data can be unknowingly violated, leading to unexpected outcome. Our approach advocates for making these assumptions -- often simple or taken for granted -- explicit, in order to more efficiently diagnose unexpected outcomes. **

2. The proposed procedure appears to assume that it is possible to simulate datasets similar to the observed data (e.g., as illustrated in Figure 2). However, this assumption may not be realistic in many practical scenarios. Simulating data requires a data-generating model, but commonly used statistical methods such as generalized estimating equations (GEE) or the generalized method of moments (GMM) do not fully specify the underlying distribution or likelihood. Even when the mean structure is specified, there may be infinitely many likelihood models consistent with it. Simulating from only one such model risks overlooking a broad class of plausible alternatives. How do the authors propose to handle such cases? More broadly, generating synthetic data with properties similar to real data is a fundamental and challenging problem in data science practice. The authors may wish to comment on recent advances in this area that aim to address this challenge without fully specifying a probabilistic model. **A: The reviewer is correct that in certain scenarios (e.g. GEE and GMM), it can be challenging to accurately simulate dataset that resemble the observed data. However, the primary aim of our paper is not to recover the true data generating process or to find the model that best fits the data. Instead, our goal is to understand which characteristic of the data may lead to the unexpected results - a mismatch between the data and the model. Here, we hold the model fixed and examine the idiosyncrasies in the data that could causes such incompatibility. This contrasts with the approach of treating data as fixed and focus on improving the model. Our perspective is motivated by the applied studies, where certain modelling strategies are generally accepted for specific types of data. When unexpected outcomes arise, analysts often first investigate whether particular characteristics of the data may cause the results, rather than developing new models. In this context, the analysis validation checks we propose provide a pre-developed "diagnostic checks" to investigate the data. To note that, even when the results is as expected, we don't necessarily assume the model is correct. [More to address the second part]**

<!-- Train AI model with synthetic data. Take the image and rotate 45 degree. - maybe talk about it in the discussion** -->

3. The construction of validation checks appears to be a critical step in applying the proposed methodology. It would be valuable if the authors could offer some guidance or recommendations on the principles for designing these checks. Clarifying this aspect would enhance the practical utility of the approach and help readers apply it more effectively in real-world settings.

**A: In the toy example section (Section 3.1), we include some explanations on the construction of validation checks: these are diagnostics analysts may naturally check when they receive an unexpected outcome. For example, when the mean of a vector falls outside an expected range, analysts may check the the five-number summary or plot the variable distribution. The validation checks are those diagnostics framed as binary predicates. **

4. The independence metric appears to be ill-defined or uninformative for single checks, much like defining total correlation for a single variable. By definition, the metric seems to always take the value one in such cases, as observed in Tables 1 and 3. Including it in the total score may artificially inflate the overall score when compared to checks involving multiple components. Further, in the current examples, the multivariate checks consistently yield independence values close to one. Are there cases where this metric meaningfully deviates from one? Clarifying the interpretation and role of this metric, particularly in the single-check case, would strengthen the methodology. **A: I need to investigate on this.**

5. The writing is generally clear and easy to follow. However, there are a number of minor syntax errors that could be addressed with more careful proofreading. The use of AI-based writing assistance tools could be helpful in this regard and, in my view, would constitute an ethical and constructive application of such tools. **A: We have run though the spelling check on the article and fixed the spelling mistakes.**

@@

## Specific Comments

1.  Section 4, paragraph 2: Should the “simulated datasets” be independent of the observed data? If so, this is a big difference from bootstrap samples and a remark would be useful. **A: The simulated dataset in our paper are not necessarily independent of the observed data, and we do not rely on bootstrap sampling at any point, although it can be a method to generate the simulated data. We have include a discussion of the principle and different strategies to generate the simulated data after paragraph 2 in Section 4.**

2.  Figure 2: Quadratic mean is mentioned in the text but not shown in the SCORE box. **A: easy. TODO**

3.  Section 4.1, paragraph 1: X1, . . . , Xn’s are undefined at the first occurrence; n means the number of checks, but in statistics it is often used for sample size. **A: We have added the definition of $X_i$ as the binary predictors in Section 4.1 and replace $n$ with $k$ for the number of checks. **

@@

4.  Section 4.2, end of paragraph 1: I assume the dash should be the minus sign. Having a minus sign in a sentence is confusing. This needs to be rephrased, perhaps just use “minus”.  **A: We have replaced the dash with the word "minus" in the text. **

5.  Section 4.2, paragraph 2: What is the range of total correlation? My understanding is that it is \[0, ∞). Even after standardizing to per observation, the range may still be not \[0, 1) as desired. **A: The range of total correlation is \[0, ∞), to obtain a measure in \[0, 1), we calculate total correlation per observation as the independence score. This is explained in the last sentence of the paragraph. TODO**

6.  Continue: p is undefined; the equation needs to close with a period.  **A: We have added the definition of p(.) and close the sentence.**

7.  Table 2: How did the authors come up with these checks? Readers would appreciate the motivations of these checks. Some of the checks are apparently correlated (e.g., the four checks on mortality-temperature correlation). **A: Four mortality-pm10/ mortality-temperature tests are constructed because analysts may not be sure what the best cutoff point for the correlation tests would be. This example illustrates how the method developed can be used to determine the cutoff value for tests.**

8.  Section 5.1, paragraph 1: Gaussian copula does not allow tail dependence. So by imposing a Gaussian copula, it is assumed that there is no tail dependence in these three variables. As questioned in one of my general comments, this might not be desirable. **A: ask Roger** may or may not being able to simulate a particular characteristic of the data and generate

9.  Figure 4: Could add bands of 95% confidence intervals of the QQ-plot; the texts on top of each panel could be moved to the caption; the labels of the axis should clarify empirical versus theoretical. **A: We have added the 95% confidence interval, remove the plot title to the caption, and refine the axis labels.**

@@

10. Figure 5 caption: The negative sign in front of 0.03 needs to be fixed. **A: Fixed.**

11. Section 5, paragraph 1: The “our second example” appears to be a continuation of the same analysis. **A: We use the wording "second example" because we count the toy example as the first example. But, to avoid this confusion, we have changed the wording to "In the following example".**

12. Table 4: There are four checks for the mortality-temperature correlation, but only one check reported in this table. **A: addressed in *Specific comments* #7. **

13. Section 5: It’d be valuable to see some real improvements from the checks. For example, fit a poor model first and identify the issues from the checks; then fit a strong model to fix the issues and pass the checks. **A: The reviewer's comment assumes we need a better model when an unexpected outcome occurs, however, in practice, we always assume there is something issue with the data, rather than rushing into developing new methods. The analysis validation checks are used to motivate a better understanding of the data, not an ideal model (This point is also addressed in *General comments* #2).**

### 2025-07-20 _manuscript/_tex/index.tex
more responses

We might gain insight into analysts' thought processes by speaking with
them directly or watching them work via screencast videos they produce,
such as, TidyTuesday screencast videos or think-aloud type studies
\citep[e.g.][]{gu2024data}. However, direct observation of analysis is
not scalable and may not always be feasible; creating educational
screencast videos requires significant effort from the analysts.
Ideally, there could be a way to explicitly communicate these thought
processes in the data analysis to others. Even better, if these
expectations were machine-readable, we could analyze them and learn from
the analysis itself. For example, we could answer questions about
whether the checks also apply to other researchers analyzing new data in
the same context, whether they reflect common practices in the field, or
whether they are specific to the data or analysis at hand. Externalizing
these thought processes is a practice that has potential to improve the
trustworthiness of the analyses \citep{yu2024veridical} in general and
the trustworthiness of subsequent products, such as machine learning
models, that may be built on such analyses. Developing an approach to
reveal some of this process without requiring a reader to essentially
reconstruct the analysis process from the raw data would provide an
improved basis for trusting an analysis result
\citep{peng2021reproducible}.

@@

To diagnose potential reasons why this expectation might fail, we can
establish a few analysis validation checks in anticipation of seeing the
data. For example, we can check the quantile of the step count, if more
than a third of the days fall below 8,000, or more than a third exceed
10,000 steps, this could indicate an excess of low-count or high-count
days. Similarly, we may may expect the standard deviation of the step
count not to be overly large. These considerations yield the following
three analysis validation checks that \emph{fail} when:

@@

The approach uses simulated datasets that have the same structure and
characteristics of the observed data to be analyzed in the future. For
each simulated dataset, analysis validation checks are applied to record
their TRUE or FALSE values. The outcome of the analysis is also computed
to determine whether it is unexpected (1) or expected (0). By generating
many datasets, each time applying the analysis validation checks and
computing the outcome, we can relate the patterns in the checks to the
likely that the outcome will be unexpected. Figure~\ref{fig-metric-calc}
provides an overview of the process.

@@

From the simulated data, the accuracy branch in
Figure~\ref{fig-metric-calc} refers to the set of checks' ability to
detect unexpected outcomes accurately. To relate multiple checks to the
outcome, a logic regression model \citep{ruczinski_logic_2003} is used.
Originally developed for SNP micro-array data, logic regression
constructs Boolean combinations of binary variables,
\(\mathcal{L}(X_1, X_2, \cdots, X_n)\), to predict classification or
regression outcomes. The logical operations allowed to construct
\(\mathcal{L(\cdot)}\) includes AND (\(\land\)), OR (\(\lor\)), and NOT
(\(\neg\)) and an example of the Boolean combination of
\(\mathcal{L(\cdot)}\) is \(X_1 \text{ AND } (X_2 \text{ OR } X_3)\). In
our binary-binary classification problem, the logic regression model can
be written as follows:

\[E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_n)\]

where the outcome, \(Y\), indicates whether the outcome is
\emph{unexpected} (1) or \emph{expected} (0), and the analysis
validation checks, \(X_1, X_2, \cdots, X_n\), are labeled as 1 if they
fail and 0 if they pass.

@@

Total correlation is a generalization of mutual information that
measures the amount of shared information across multiple variables
multiple variables based on Kullback-Leibler distance:

\[C(X_1, X_2, \cdots, X_n) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_n} p(x_1, x_2, \cdots, x_n) \log \frac{p(x_1, x_2, \cdots, x_n)}{p(x_1)p(x_2) \cdots p(x_n)},\]

where \(p(.)\) is the TODO. A high total correlation value indicates
overlapping of information among the analysis validation checks, while a
low value suggests that the checks are independent and provides unique
information to diagnose the unexpected outcome. To standardize this
measure, the total correlation \emph{per observation} is used in
calculating the independence score.

@@

\caption{\label{fig-dist-fit}QQ-plot of the distribution fit for
mortality, PM10, and temperature based on the fitted distribution from
the original data. The fitted distribution is compared to the observed
data to assess the distribution fit.}

>>>>

We might gain insight into analysts' thought processes by speaking with
them directly or watching them work via screencast videos they produce,
such as TidyTuesday screencast videos or think-aloud type studies
\citep[e.g.][]{gu2024data}. However, direct observation of analysis is
not scalable and may not always be feasible; creating educational
screencast videos requires significant effort from the analysts.
Ideally, there could be a way to explicitly communicate these thought
processes in the data analysis to others. Even better, if these
expectations were machine-readable, we could analyze them and learn from
the analysis itself. For example, we could answer questions about
whether the checks also apply to other researchers analyzing new data in
the same context, whether they reflect common practices in the field, or
whether they are specific to the data or analysis at hand. Externalizing
these thought processes can improve the trustworthiness of the analyses
\citep{yu2024veridical}.

The trustworthiness of an analysis depends not only on the robustness of
the methods used but also on a deep understanding of the data. This
corresponds to the third step identified in
\citet{broderick_toward_2023}, where trust can break down if the
assumptions underlying an algorithm are not satisfied in practice. In
the past, data were often collected directly by analysts through
carefully designed experiments, however, contemporary analyses often
rely on data curated by government or research institutions, where some
data characteristics may be unknown to the analysts. When unexpected
outcomes arise, they often later traced back to implicit assumptions
made by the analysts - assumptions that were unknowingly violated by the
idiosyncratic in the data. By externalizing these assumptions through
analysis validation checks, we aim to make it easier to identify and
diagnose such unexpected results.

@@

To diagnose potential reasons why this expectation might fail, we can
establish a few analysis validation checks in anticipation of seeing the
data. {[}TODO: Add guidelines for designing the checks{]} For example,
we can check the quantile of the step count, if more than a third of the
days fall below 8,000, or more than a third exceed 10,000 steps, this
could indicate an excess of low-count or high-count days. Similarly, we
may may expect the standard deviation of the step count not to be overly
large. These considerations yield the following three analysis
validation checks that \emph{fail} when:

@@

The approach generates simulated datasets from the observed data to
compute the outcome and the analysis validation checks (unexpected as 1
and expected as 0). This then allows us to apply a logic regression tree
to link the outcome and the check results to construct a Boolean
expression to best predict the unexpected outcomes.
Figure~\ref{fig-metric-calc} provides an overview of the process.

The specific strategy to simulate the data depends on the problem and
the nature of the data at hand. Simulated data can be generated using
prior domain knowledge, summary statistics derived from the observed
data, or statistical techniques such as training-testing splits or
bootstrap sampling. While each simulated data may not be independent of
the original data - as is the case with bootstrap samples - it should
adequately capture the range of possible idiosyncrasies present in the
data. Failure to do so can reduce the ability of the resulting Boolean
expression to predict unexpected outcomes.

@@

From the simulated data, the accuracy branch in
Figure~\ref{fig-metric-calc} refers to the set of checks' ability to
detect unexpected outcomes accurately. To relate multiple checks to the
outcome, a logic regression model \citep{ruczinski_logic_2003} is used.
Originally developed for SNP micro-array data, logic regression predict
classification or regression outcomes by constructing Boolean
combinations, \(\mathcal{L}(X_1, X_2, \cdots, X_k)\), where \(X_i\) are
binary predictors. The logical operations allowed to construct
\(\mathcal{L(\cdot)}\) includes AND (\(\land\)), OR (\(\lor\)), and NOT
(\(\neg\)) and an example of the Boolean combination of
\(\mathcal{L(\cdot)}\) is \(X_1 \text{ AND } (X_2 \text{ OR } X_3)\). In
our binary-binary classification problem, the logic regression model can
be written as follows:

\[E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_k)\]

where the outcome, \(Y\), indicates whether the outcome is
\emph{unexpected} (1) or \emph{expected} (0), and the analysis
validation checks, \(X_1, X_2, \cdots, X_k\), are labeled as 1 if they
fail and 0 if they pass.

@@

Total correlation is a generalization of mutual information that
measures the amount of shared information across multiple variables,
based on Kullback-Leibler distance:

\[C(X_1, X_2, \cdots, X_k) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_k} p(x_1, x_2, \cdots, x_k) \log \frac{p(x_1, x_2, \cdots, x_k)}{p(x_1)p(x_2) \cdots p(x_k)},\]

where \(p(x_1, x_2, \cdots, x_k)\) is the joint distribution of the
predictors and \(p(\cdot)\) is the individual distribution of the
predictors. A high total correlation value indicates overlapping of
information among the analysis validation checks, while a low value
suggests that the checks are independent and provides unique information
to diagnose the unexpected outcome. To standardize this measure, the
total correlation \emph{per observation} is used in calculating the
independence score.

@@

\caption{\label{fig-dist-fit}QQ-plot of the goodness-of-fit of selected
distributions used to generate simulated data for mortality, PM10, and
temperature. Left: Mortality as a negative binomial distribution (size =
74, mean = 183), Middle: PM10 is modeled using a rescaled beta
distribution with shape parameters 4.21 and 11.67, multiplied by 10 to
match the observed range. Right: Temperature is modeled using a Weibull
distribution (shape = 3.8, scale = 61)}

### 2025-07-20 _manuscript/index.qmd
more responses

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
#remotes::install_github("huizezhang-sherry/adtoolbox")
library(tidyverse)
library(patchwork)
library(MASS)
library(ggh4x)
library(adtoolbox)
library(broom)
library(knitr)
library(rpart)
library(infotheo)
library(patchwork)
library(readr)
library(kableExtra)
library(scales)
pm10 <- read_csv(here::here("data", "pm10.csv")) |> 
  filter(!is.na(pm10)) |>    
  mutate(season = as.factor(season))  
```

@@

We might gain insight into analysts' thought processes by speaking with them directly or watching them work via screencast videos they produce, such as, TidyTuesday screencast videos or think-aloud type studies [e.g. @gu2024data]. However, direct observation of analysis is not scalable and may not always be feasible; creating educational screencast videos requires significant effort from the analysts. Ideally, there could be a way to explicitly communicate these thought processes in the data analysis to others. Even better, if these expectations were machine-readable, we could analyze them and learn from the analysis itself. For example, we could answer questions about whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect common practices in the field, or whether they are specific to the data or analysis at hand. Externalizing these thought processes is a practice that has potential to improve the trustworthiness of the analyses [@yu2024veridical] in general and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses. Developing an approach to reveal some of this process without requiring a reader to essentially reconstruct the analysis process from the raw data would provide an improved basis for trusting an analysis result [@peng2021reproducible].

@@

To diagnose potential reasons why this expectation might fail, we can establish a few analysis validation checks in anticipation of seeing the data. For example, we can check the quantile of the step count, if more than a third of the days fall below 8,000, or more than a third exceed 10,000 steps, this could indicate an excess of low-count or high-count days. Similarly, we may may expect the standard deviation of the step count not to be overly large. These considerations yield the following three analysis validation checks that *fail* when:

@@

The approach uses simulated datasets that have the same structure and characteristics of the observed data to be analyzed in the future. For each simulated dataset, analysis validation checks are applied to record  their TRUE or FALSE values. The outcome of the analysis is also computed to determine whether it is unexpected (1) or expected (0). By generating many datasets, each time applying the analysis validation checks and computing the outcome, we can relate the patterns in the checks to the likely that the outcome will be unexpected. @fig-metric-calc provides an overview of the process.

@@

From the simulated data, the accuracy branch in @fig-metric-calc refers to the set of checks' ability to detect unexpected outcomes accurately. To relate multiple checks to the outcome, a logic regression model [@ruczinski_logic_2003] is used. Originally developed for SNP micro-array data, logic regression constructs Boolean combinations of binary variables, $\mathcal{L}(X_1, X_2, \cdots, X_n)$, to predict classification or regression outcomes. The logical operations allowed to construct $\mathcal{L(\cdot)}$ includes AND ($\land$), OR ($\lor$), and NOT ($\neg$) and an example of the Boolean combination of $\mathcal{L(\cdot)}$ is $X_1 \text{ AND } (X_2 \text{ OR } X_3)$. In our binary-binary classification problem, the logic regression model can be written as follows:

$$E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_n)$$

where the outcome, $Y$, indicates whether the outcome is *unexpected* (1) or *expected* (0), and the analysis validation checks, $X_1, X_2, \cdots, X_n$, are labeled as 1 if they fail and 0 if they pass.

@@

Total correlation is a generalization of mutual information that measures the amount of shared information across multiple variables multiple variables based on Kullback-Leibler distance:

$$C(X_1, X_2, \cdots, X_n) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_n} p(x_1, x_2, \cdots, x_n) \log \frac{p(x_1, x_2, \cdots, x_n)}{p(x_1)p(x_2) \cdots p(x_n)},$$

where $p(.)$ is the TODO. A high total correlation value indicates overlapping of information among the analysis validation checks, while a low value suggests that the checks are independent and provides unique information to diagnose the unexpected outcome. To standardize this measure, the total correlation *per observation* is used in calculating the independence score.

@@

```{r}
#| label: fig-dist-fit
#| fig-cap: QQ-plot of the distribution fit for mortality, PM10, and temperature based on the fitted distribution from the original data. The fitted distribution is compared to the observed data to assess the distribution fit. 
#| fig.height: 4
#| fig.width: 10
aa2 <- fitdistrplus::fitdist(pm10$mortality, "nbinom")
p1 <- tibble(fitted= rnbinom(558, size = aa2$estimate[1], mu = aa2$estimate[2]),
       observed= pm10$mortality) %>%
  ggplot(aes(sample = fitted)) +
  stat_qq() +
  stat_qq_line() + 
  theme_bw() + 
  theme(aspect.ratio = 1) + 
  ggtitle("mortality: \nNB(size = 74, mu = 183)")

aa6 <- fitdistrplus::fitdist(pm10$pm10[pm10$pm10 < 90]/100, "beta")
p2 <- tibble(fitted = rbeta(557, shape1 = aa6$estimate[1], shape2 = aa6$estimate[2]) * 100,
       observed = pm10$pm10[pm10$pm10 < 90]) %>%
  ggplot(aes(sample = fitted)) +
  stat_qq() +
  stat_qq_line() + 
  theme_bw() + 
  theme(aspect.ratio = 1) + 
  ggtitle("pm10: \n100 * beta(shp1 = 4.21, shp2 = 11.67)")

aa4 <- fitdistrplus::fitdist(pm10$temp, "weibull")
p3 <- tibble(fitted= rweibull(558, aa4$estimate[1], aa4$estimate[2]),
       observed= pm10$temp) %>%
  ggplot(aes(sample = fitted)) +
  stat_qq() +
  stat_qq_line() + 
  theme_bw() + 
  theme(aspect.ratio = 1) + 
  ggtitle("temperature: \nweibull(shape = 3.8, scale = 61)")

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
#remotes::install_github("huizezhang-sherry/adtoolbox")
library(tidyverse)
library(patchwork)
library(MASS)
library(ggh4x)
library(adtoolbox)
library(broom)
library(knitr)
library(rpart)
library(infotheo)
library(patchwork)
library(readr)
library(kableExtra)
library(scales)
library(qqplotr)
pm10 <- read_csv(here::here("data", "pm10.csv")) |> 
  filter(!is.na(pm10)) |>    
  mutate(season = as.factor(season))  
```

@@

We might gain insight into analysts' thought processes by speaking with them directly or watching them work via screencast videos they produce, such as TidyTuesday screencast videos or think-aloud type studies [e.g. @gu2024data]. However, direct observation of analysis is not scalable and may not always be feasible; creating educational screencast videos requires significant effort from the analysts. Ideally, there could be a way to explicitly communicate these thought processes in the data analysis to others. Even better, if these expectations were machine-readable, we could analyze them and learn from the analysis itself. For example, we could answer questions about whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect common practices in the field, or whether they are specific to the data or analysis at hand. Externalizing these thought processes can improve the trustworthiness of the analyses [@yu2024veridical].

The trustworthiness of an analysis depends not only on the robustness of the methods used but also on a deep understanding of the data. This corresponds to the third step identified in @broderick_toward_2023, where trust can break down if the assumptions underlying an algorithm are not satisfied in practice. In the past, data were often collected directly by analysts through carefully designed experiments, however, contemporary analyses often rely on data curated by government or research institutions, where some data characteristics may be unknown to the analysts. When unexpected outcomes arise, they often later traced back to implicit assumptions made by the analysts - assumptions that were unknowingly violated by the idiosyncratic in the data. By externalizing these assumptions through analysis validation checks, we aim to make it easier to identify and diagnose such unexpected results.

@@

To diagnose potential reasons why this expectation might fail, we can establish a few analysis validation checks in anticipation of seeing the data. [TODO: Add guidelines for designing the checks] For example, we can check the quantile of the step count, if more than a third of the days fall below 8,000, or more than a third exceed 10,000 steps, this could indicate an excess of low-count or high-count days. Similarly, we may may expect the standard deviation of the step count not to be overly large. These considerations yield the following three analysis validation checks that *fail* when:

@@

The approach generates simulated datasets from the observed data to compute the outcome and the analysis validation checks (unexpected as 1 and expected as 0). This then allows us to apply a logic regression tree to link the outcome and the check results to construct a Boolean expression to best predict the unexpected outcomes. @fig-metric-calc provides an overview of the process.

The specific strategy to simulate the data depends on the problem and the nature of the data at hand. Simulated data can be generated using prior domain knowledge, summary statistics derived from the observed data, or statistical techniques such as training-testing splits or bootstrap sampling. While each simulated data may not be independent of the original data - as is the case with bootstrap samples - it should adequately capture the range of possible idiosyncrasies present in the data. Failure to do so can reduce the ability of the resulting Boolean expression to predict unexpected outcomes.

@@

From the simulated data, the accuracy branch in @fig-metric-calc refers to the set of checks' ability to detect unexpected outcomes accurately. To relate multiple checks to the outcome, a logic regression model [@ruczinski_logic_2003] is used. Originally developed for SNP micro-array data, logic regression predict classification or regression outcomes by constructing Boolean combinations, $\mathcal{L}(X_1, X_2, \cdots, X_k)$, where $X_i$ are binary predictors. The logical operations allowed to construct $\mathcal{L(\cdot)}$ includes AND ($\land$), OR ($\lor$), and NOT ($\neg$) and an example of the Boolean combination of $\mathcal{L(\cdot)}$ is $X_1 \text{ AND } (X_2 \text{ OR } X_3)$. In our binary-binary classification problem, the logic regression model can be written as follows:

$$E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_k)$$

where the outcome, $Y$, indicates whether the outcome is *unexpected* (1) or *expected* (0), and the analysis validation checks, $X_1, X_2, \cdots, X_k$, are labeled as 1 if they fail and 0 if they pass.

@@

Total correlation is a generalization of mutual information that measures the amount of shared information across multiple variables, based on Kullback-Leibler distance:

$$C(X_1, X_2, \cdots, X_k) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_k} p(x_1, x_2, \cdots, x_k) \log \frac{p(x_1, x_2, \cdots, x_k)}{p(x_1)p(x_2) \cdots p(x_k)},$$

where $p(x_1, x_2, \cdots, x_k)$ is the joint distribution of the predictors and $p(\cdot)$ is the individual distribution of the predictors. A high total correlation value indicates overlapping of information among the analysis validation checks, while a low value suggests that the checks are independent and provides unique information to diagnose the unexpected outcome. To standardize this measure, the total correlation *per observation* is used in calculating the independence score.

@@

```{r}
#| label: fig-dist-fit
#| fig-cap: "QQ-plot of the goodness-of-fit of selected distributions used to generate simulated data for mortality, PM10, and temperature. Left: Mortality as a negative binomial distribution (size = 74, mean = 183), Middle: PM10 is modeled using a rescaled beta distribution with shape parameters 4.21 and 11.67, multiplied by 10 to match the observed range. Right: Temperature is modeled using a Weibull distribution (shape = 3.8, scale = 61)"
#| fig.height: 4
#| fig.width: 10
aa2 <- fitdistrplus::fitdist(pm10$mortality, "nbinom")
p1 <- tibble(fitted = rnbinom(558, size = aa2$estimate[1], mu = aa2$estimate[2]),
       observed = pm10$mortality) %>%
  ggplot(aes(sample = fitted)) +
  stat_qq_band(fill = "grey80") +
  stat_qq_line(color = "grey30") +
  stat_qq_point(size = 0.3) +
  theme_bw() + 
  theme(aspect.ratio = 1) + 
  ggtitle("Mortality")

aa6 <- fitdistrplus::fitdist(pm10$pm10[pm10$pm10 < 90]/100, "beta")
p2 <- tibble(fitted = rbeta(557, shape1 = aa6$estimate[1], shape2 = aa6$estimate[2]) * 100,
       observed = pm10$pm10[pm10$pm10 < 90]) %>%
  ggplot(aes(sample = fitted)) +
  stat_qq_band(fill = "grey80") +
  stat_qq_line(color = "grey30") +
  stat_qq_point(size = 0.3) +
  theme_bw() + 
  theme(aspect.ratio = 1) +
  ggtitle("PM10")

aa4 <- fitdistrplus::fitdist(pm10$temp, "weibull")
p3 <- tibble(fitted= rweibull(558, aa4$estimate[1], aa4$estimate[2]),
       observed= pm10$temp) %>%
  ggplot(aes(sample = fitted)) +
  stat_qq_band(fill = "grey80") +
  stat_qq_line(color = "grey30") +
  stat_qq_point(size = 0.3) +
  theme_bw() + 
  theme(aspect.ratio = 1) +
  ggtitle("Temperature")
